The effects of supervision upon effort during resistance training: A Bayesian analysis of prior data and an experimental study of private strength clinic members

### 2025-03-04 submission/pre_print_jss_submission.qmd
Update preprint, and add journal submission outputs

---
articletitle: |
  The effects of supervision upon effort during resistance training: A Bayesian analysis of prior data and an experimental study of private strength clinic members
# shorttitle: "Doe et al. (2022)" # usually auto-generated
format: 
  sportrxiv-pdf:
    include-in-header:
      text: |
       \usepackage[font=scriptsize]{caption}
       \usepackage{caption}
author:
  - name: James Steele
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 3
    orcid: 0000-0002-8003-0757
    corresponding: true
    email: james@steele-research.com
  - name: James Fisher
    affiliations:
      - ref: 3
      - ref: 4
    orcid: 0000-0002-6013-8402
    email: james.fisher.phd@outlook.com
  - name: Tim Dettmann
    affiliations:
      - ref: 1
    email: tim.dettmann@gmail.com
affiliations:
      - id: 1
        name: Kieser Australia, Melbourne, Australia
      - id: 2
        name: Steele Research Limited, Eastleigh, UK
      - id: 3
        name: Department of Sport and Health, Solent University, Southampton, UK
      - id: 4
        name: The Exercise Coach, Tennessee, USA
abstract: |
  Supervision during resistance training (RT) may enhance strength gains by optimizing trainee effort. We investigated supervision’s role in effort during RT in a unique setting with private strength clinics, where members train either unsupervised (“Core” membership) or supervised by a qualified exercise scientist (“Assisted” membership). Using both retrospective analysis of member training records and a prospective experimental study, we examined supervision’s impact on exercise performance, measured as time under load (TUL), rating of perceived effort (RPE), and rating of perceived discomfort (RPD). Bayesian methods were applied, using empirically informed prior distributions from retrospective data to model the experimental study. The prior sample included ~1000 members training sessions from each membership type, while the experimental study involved 45 Core members performing both supervised and unsupervised sessions in randomized order, using their current training loads to momentary failure. Our findings suggest that, in real-world settings (in situ), exercise performance differed little between supervised and unsupervised training. However, in our experimental study, supervision improved TUL (Core = 125.12 [95%QI: 113.70, 131.90] sec; Assisted = 147.35 [95%QI: 134.29, 154.81] sec; contrast = -22.10 [95%QI: -26.60, -17.61] sec). In percentage points RPE was slightly higher with supervision in both prior real-world (Core = 53% [95%QI: 51%, 55%]; Assisted = 59% [95%QI: 57%, 61%]; contrast = -6% [95%QI: -8%, -4%]) and experimental settings (Core = 81% [95%QI: 75%, 86%]; Assisted = 87% [95%QI: 83%, 91%]; contrast = -6% [95%QI: -10%, -4%]), suggesting trainees push closer to failure under supervision. This was further supported by higher RPD during the experimental study (Core = 6.3 [95%QI: 5.1, 7.3]; Assisted = 7.5 [95%QI: 6.5, 8.3]; contrast = -1.2 [95%QI: -1.6, -0.9]). Overall, these results reinforce prior research on the benefits of supervision in RT, indicating that unsupervised trainees—especially in real-world conditions—likely train with suboptimal effort.
license-type: ccby # change if neccessary
# year: 2025 # defaults to current year
keywords: [supervision, effort, resistance training, bayesian] # optional
# optional link to Supplementary Material:
suppl-link: https://osf.io/etb34/
reference-section-title: References
printnote: "PREPRINT - NOT PEER REVIEWED" # if post print, include "POSTPRINT" then link to the published article
bibliography: bibliography.bib  
pdf-engine: xelatex
execute: 
  echo: false
  message: false
  warning: false
---

@@

@@

$$
\tiny
\begin{aligned}
\text{RPE}_{ic} &\sim
\begin{Bmatrix}
0 & \text{with probability } \alpha_{ic_{j[i],k[i],l[i]}} \\[4pt]
\in (0,1) & \text{with probability } \delta_{ic_{j[i],k[i],l[i]}} \\[4pt]
1 & \text{with probability } \gamma_{ic_{j[i],k[i],l[i]}} \\[4pt]
\end{Bmatrix}
\\
\\
\text{logit}(\alpha_{ic}) &= 1-(X'\beta-k_1) & \text{Probability of obtaining a 0} \\[4pt]
\text{logit}(\delta_{ic}) &= \bigr[(X'\beta-k_1) - (X'\beta-k_2)\bigr]\text{Beta}(X'\beta, X'
\beta_\phi) & \text{Probability of obtaining a value between 0 and 1} \\[4pt]
\text{logit}(\gamma_{ic}) &= (X'\beta-k_2) & \text{Probability of obtaining a 1} \\[4pt]
\\
X'\beta&=(\beta_0 + b_{0_{j[i],k[i],l[i]}}) + \beta_1 \text{Condition}_{ic} & \text{Vector of predictors} \\[4pt] 
b_{0_j} &\sim \mathcal{N}(0, \sigma_{b_{0_{j}}}), \text{for location j} = 1,\ ...,\text{J} & \text{Variation in intercepts for location in student}\ t\ \text{model model} \\[4pt]
b_{0_k} &\sim \mathcal{N}(0, \sigma_{b_{0_{k}}}), \text{for member k} = 1,\ ...,\text{K} & \text{Variation in intercepts for member in student}\ t\ \text{model model} \\[4pt]
b_{0_l} &\sim \mathcal{N}(0, \sigma_{b_{0_{l}}}), \text{for machine l} = 1,\ ...,\text{L} & \text{Variation in intercepts for machine  in student}\ t\ \text{model model} \\[4pt]
X'\beta_\phi&=(\beta_{\phi0} + b_{\phi0_{j[i],k[i],l[i]}}) + \beta_{\phi1} \text{Condition}_{ic} & \text{Vector of predictors} \\[4pt] 
b_{\phi0_j} &\sim \mathcal{N}(0, \sigma_{b_{\phi0_{j}}}), \text{for location j} = 1,\ ...,\text{J} & \text{Variation in intercepts for location in student}\ t\ \text{model model} \\[4pt]
b_{\phi0_k} &\sim \mathcal{N}(0, \sigma_{b_{\phi0_{k}}}), \text{for member k} = 1,\ ...,\text{K} & \text{Variation in intercepts for member in student}\ t\ \text{model model} \\[4pt]
b_{\phi0_l} &\sim \mathcal{N}(0, \sigma_{b_{\phi0_{l}}}), \text{for machine l} = 1,\ ...,\text{L} & \text{Variation in intercepts for machine  in student}\ t\ \text{model model} \\[4pt]
\\
\end{aligned}
$$ {#eq-ordbeta-model}

@@

>>>>

---
title: "The effects of supervision upon effort during resistance training: A Bayesian analysis of prior data and an experimental study of private strength clinic members"
author:
  - name: James Steele
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 3
    orcid: 0000-0002-8003-0757
    corresponding: true
    email: james@steele-research.com
  - name: James Fisher
    affiliations:
      - ref: 3
      - ref: 4
    orcid: 0000-0002-6013-8402
    email: james.fisher.phd@outlook.com
  - name: Tim Dettmann
    affiliations:
      - ref: 1
    email: tim.dettmann@gmail.com
affiliations:
      - id: 1
        name: Kieser Australia, Melbourne, Australia
      - id: 2
        name: Steele Research Limited, Eastleigh, UK
      - id: 3
        name: Department of Sport and Health, Solent University, Southampton, UK
      - id: 4
        name: The Exercise Coach, Tennessee, USA
abstract: |
  Supervision during resistance training (RT) may optimise trainee effort. We examined this in private strength clinics, where members train either unsupervised (“Core”) or supervised by an exercise scientist (“Assisted”). Using retrospective analysis of real-world data from ~1000 members training sessions per group, and an experimental study with 45 Core members, we assessed supervision’s impact on time under load (TUL), and rating of perceived effort (RPE) and discomfort (RPD). Bayesian methods were used to integrate retrospective data priors and update to posteriors using data from the experimental study. Real-world retrospective data showed minimal performance differences, but the experimental study posterior estimates found supervision improved TUL (Core = 125.12 [95%QI: 113.70, 131.90] sec; Assisted = 147.35 [95%QI: 134.29, 154.81] sec; contrast = -22.10 [95%QI: -26.60, -17.61] sec). Supervision also slightly increased RPE in real-world (Core = 53% [95%QI: 51%, 55%]; Assisted = 59% [95%QI: 57%, 61%]; contrast = -6% [95%QI: -8%, -4%]) and experimental settings (Core = 81% [95%QI: 75%, 86%]; Assisted = 87% [95%QI: 83%, 91%]; contrast = -6% [95%QI: -10%, -4%])), indicating trainees pushed closer to failure. RPD was also higher with supervision. These findings reinforce prior research on supervision’s benefits, suggesting unsupervised trainees—especially in real-world conditions—train with suboptimal effort.
format:
  pdf: 
    include-in-header:
      text: '\usepackage{lineno}\linenumbers'
bibliography: bibliography.bib  
pdf-engine: xelatex
execute: 
  echo: false
  message: false
  warning: false
---

@@

## Author Affiliations
James Steele: Kieser Australia, Melbourne, Australia; Steele Research Limited, Eastleigh, UK; Department of Sport and Health, Solent University, Southampton, UK
James Fisher: Department of Sport and Health, Solent University, Southampton, UK; The Exercise Coach, Tennessee, USA
Tim Dettmann: Kieser Australia, Melbourne, Australia

Corresponding author: James Steele (james@steele-research.com)

@@

$$
\tiny
\begin{aligned}
\text{RPE}_{ic} &\sim
\begin{Bmatrix}
0 & \text{with probability } \alpha_{ic_{j[i],k[i],l[i]}} \\[4pt]
\in (0,1) & \text{with probability } \delta_{ic_{j[i],k[i],l[i]}} \\[4pt]
1 & \text{with probability } \gamma_{ic_{j[i],k[i],l[i]}} \\[4pt]
\end{Bmatrix}
\\
\\
\text{logit}(\alpha_{ic}) &= 1-(X'\beta-k_1) & \text{Probability of obtaining a 0} \\[4pt]
\text{logit}(\delta_{ic}) &= \bigr[(X'\beta-k_1) - (X'\beta-k_2)\bigr]\text{Beta}(X'\beta, X'
\beta_\phi) & \text{Probability of obtaining a value between 0 and 1} \\[4pt]
\text{logit}(\gamma_{ic}) &= (X'\beta-k_2) & \text{Probability of obtaining a 1} \\[4pt]
\\
X'\beta&=(\beta_0 + b_{0_{j[i],k[i],l[i]}}) + \beta_1 \text{Condition}_{ic} & \text{Vector of predictors for mean parameter} \\[4pt] 
b_{0_j} &\sim \mathcal{N}(0, \sigma_{b_{0_{j}}}), \text{for location j} = 1,\ ...,\text{J} & \text{Variation in intercepts for location in mean parameter} \\[4pt]
b_{0_k} &\sim \mathcal{N}(0, \sigma_{b_{0_{k}}}), \text{for member k} = 1,\ ...,\text{K} & \text{Variation in intercepts for member in mean parameter} \\[4pt]
b_{0_l} &\sim \mathcal{N}(0, \sigma_{b_{0_{l}}}), \text{for machine l} = 1,\ ...,\text{L} & \text{Variation in intercepts for machine  in mean parameter} \\[4pt]
X'\beta_\phi&=(\beta_{\phi0} + b_{\phi0_{j[i],k[i],l[i]}}) + \beta_{\phi1} \text{Condition}_{ic} & \text{Vector of predictors for precision parameter} \\[4pt] 
b_{\phi0_j} &\sim \mathcal{N}(0, \sigma_{b_{\phi0_{j}}}), \text{for location j} = 1,\ ...,\text{J} & \text{Variation in intercepts for location in precision parameter} \\[4pt]
b_{\phi0_k} &\sim \mathcal{N}(0, \sigma_{b_{\phi0_{k}}}), \text{for member k} = 1,\ ...,\text{K} & \text{Variation in intercepts for member in precision parameter} \\[4pt]
b_{\phi0_l} &\sim \mathcal{N}(0, \sigma_{b_{\phi0_{l}}}), \text{for machine l} = 1,\ ...,\text{L} & \text{Variation in intercepts for machine  in precision parameter} \\[4pt]
\\
\end{aligned}
$$ {#eq-ordbeta-model}

@@

# References

### 2025-03-04 submission/pre_print_jss_submission_anon.qmd
Update preprint, and add journal submission outputs

---
title: "The effects of supervision upon effort during resistance training: A Bayesian analysis of prior data and an experimental study of private strength clinic members"
author:
  - name: James Steele
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 3
    orcid: 0000-0002-8003-0757
    corresponding: true
    email: james@steele-research.com
  - name: James Fisher
    affiliations:
      - ref: 3
      - ref: 4
    orcid: 0000-0002-6013-8402
    email: james.fisher.phd@outlook.com
  - name: Tim Dettmann
    affiliations:
      - ref: 1
    email: tim.dettmann@gmail.com
affiliations:
      - id: 1
        name: Kieser Australia, Melbourne, Australia
      - id: 2
        name: Steele Research Limited, Eastleigh, UK
      - id: 3
        name: Department of Sport and Health, Solent University, Southampton, UK
      - id: 4
        name: The Exercise Coach, Tennessee, USA
abstract: |
  Supervision during resistance training (RT) may optimise trainee effort. We examined this in private strength clinics, where members train either unsupervised (“Core”) or supervised by an exercise scientist (“Assisted”). Using retrospective analysis of real-world data from ~1000 members training sessions per group, and an experimental study with 45 Core members, we assessed supervision’s impact on time under load (TUL), and rating of perceived effort (RPE) and discomfort (RPD). Bayesian methods were used to integrate retrospective data priors and update to posteriors using data from the experimental study. Real-world retrospective data showed minimal performance differences, but the experimental study posterior estimates found supervision improved TUL (Core = 125.12 [95%QI: 113.70, 131.90] sec; Assisted = 147.35 [95%QI: 134.29, 154.81] sec; contrast = -22.10 [95%QI: -26.60, -17.61] sec). Supervision also slightly increased RPE in real-world (Core = 53% [95%QI: 51%, 55%]; Assisted = 59% [95%QI: 57%, 61%]; contrast = -6% [95%QI: -8%, -4%]) and experimental settings (Core = 81% [95%QI: 75%, 86%]; Assisted = 87% [95%QI: 83%, 91%]; contrast = -6% [95%QI: -10%, -4%])), indicating trainees pushed closer to failure. RPD was also higher with supervision. These findings reinforce prior research on supervision’s benefits, suggesting unsupervised trainees—especially in real-world conditions—train with suboptimal effort.
format:
  pdf: 
    include-in-header:
      text: '\usepackage{lineno}\linenumbers'
bibliography: bibliography.bib  
pdf-engine: xelatex
execute: 
  echo: false
  message: false
  warning: false
---

@@

## Author Affiliations
James Steele: Kieser Australia, Melbourne, Australia; Steele Research Limited, Eastleigh, UK; Department of Sport and Health, Solent University, Southampton, UK
James Fisher: Department of Sport and Health, Solent University, Southampton, UK; The Exercise Coach, Tennessee, USA
Tim Dettmann: Kieser Australia, Melbourne, Australia

Corresponding author: James Steele (james@steele-research.com)

@@

The experimental study component of this project was approved by the Southampton Solent University Health Exercise and Sport Science Ethics committee (id: STEELEAUG2016). All participants provided informed consent to participate in the study. Data in the prior sample was used in de-identified form and with member consent for data to be used for research purposes.

@@

## Statistical Analysis
All code utilised for data preparation and analyses are available in either the Open Science Framework page for this project <https://osf.io/etb34/> or the corresponding GitHub repository <https://github.com/jamessteeleii/supervision_tul>. We cite all software and packages used in the analysis pipeline using the `grateful` package [@rodriguez-sanchezGratefulFacilitateCitation2023] which can be seen here: <https://osf.io/ew79g>.

@@

# Contributions

All authors conceived the idea for the project and designed the methods. TD coordinated and oversaw the data collection. JS conducted the statistical analyses and produced data visualisations. JS and JF drafted the initial manuscript. All authors contributed to editing the manuscript. All authors read and approved the final manuscript.

@@

All data and code utilised for data preparation and analyses are available in either the Open Science Framework page for this project https://osf.io/etb34/ or the corresponding GitHub repository https://github.com/jamessteeleii/supervision_tul. Other supplementary analyses and plots are also available there.

>>>>

---
title: "The effects of supervision upon effort during resistance training: A Bayesian analysis of prior data and an experimental study of private strength clinic members"
abstract: |
  Supervision during resistance training (RT) may optimise trainee effort. We examined this in private strength clinics, where members train either unsupervised (“Core”) or supervised by an exercise scientist (“Assisted”). Using retrospective analysis of real-world data from ~1000 members training sessions per group, and an experimental study with 45 Core members, we assessed supervision’s impact on time under load (TUL), and rating of perceived effort (RPE) and discomfort (RPD). Bayesian methods were used to integrate retrospective data priors and update to posteriors using data from the experimental study. Real-world retrospective data showed minimal performance differences, but the experimental study posterior estimates found supervision improved TUL (Core = 125.12 [95%QI: 113.70, 131.90] sec; Assisted = 147.35 [95%QI: 134.29, 154.81] sec; contrast = -22.10 [95%QI: -26.60, -17.61] sec). Supervision also slightly increased RPE in real-world (Core = 53% [95%QI: 51%, 55%]; Assisted = 59% [95%QI: 57%, 61%]; contrast = -6% [95%QI: -8%, -4%]) and experimental settings (Core = 81% [95%QI: 75%, 86%]; Assisted = 87% [95%QI: 83%, 91%]; contrast = -6% [95%QI: -10%, -4%])), indicating trainees pushed closer to failure. RPD was also higher with supervision. These findings reinforce prior research on supervision’s benefits, suggesting unsupervised trainees—especially in real-world conditions—train with suboptimal effort.
format: 
  pdf:
    include-in-header: 
      text: '\usepackage{lineno}\linenumbers'
bibliography: bibliography.bib  
pdf-engine: xelatex
execute: 
  echo: false
  message: false
  warning: false
---

@@

@@

The experimental study component of this project was approved by the a university ethics committee. All participants provided informed consent to participate in the study. Data in the prior sample was used in de-identified form and with member consent for data to be used for research purposes.

@@

## Statistical Analysis
All code utilised for data preparation and analyses are available in either the Open Science Framework page for this project <https://osf.io/etb34/>. We cite all software and packages used in the analysis pipeline using the `grateful` package [@rodriguez-sanchezGratefulFacilitateCitation2023] which can be seen here: <https://osf.io/ew79g>.

@@

@@

All data and code utilised for data preparation and analyses are available in either the Open Science Framework page for this project https://osf.io/etb34/. Other supplementary analyses and plots are also available there.
